Fast approximate Bayesian inference of HIV indicators using PCA adaptive Gauss-Hermite quadrature

### 2023-07-18 src/docs_paper/appendix.Rmd
Adding BYM2 and AR1 discussion to manuscript #55

```{r rho_a, fig.cap="Variation between units can be explained either by high correlation and high variance, or by low correlation and low variance. As such, some hyperparameters has correlated posteriors. It is this correlation which is made use of by PCA-AGHQ."}
knitr::include_graphics("depends/rho_a.png")
```

>>>>

```{r rho_a, fig.cap="Variation between units can be explained either by high correlation and high variance, or by low correlation and low variance. As such, the AR1 hyperparameters had correlated posteriors. It is this correlation which we made use of with PCA-AGHQ."}
knitr::include_graphics("depends/rho_a.png")
```

```{r rho_a, fig.cap="In contrast to the AR1 hyperparameters, the BYM2 hyperparameter posteriors are much less correlated."}
knitr::include_graphics("depends/alpha_x.png")
```

### 2023-07-18 src/docs_paper/paper.Rmd
Adding BYM2 and AR1 discussion to manuscript #55

In all, Naomi is a joint model on the observations $\mathbf{y} = (y^{\theta}_I)$ for $\theta \in \{\rho, \alpha, \kappa, \rho^\text{ANC}, \alpha^\text{ANC}, N^\text{ART}\}$ and $I \in \mathcal{I}$.
The structured additive predictors contain intercept effects, age random effects, and spatial random effects which we collectively describe as the latent field $\x$.
The latent field is controlled by hyperparamters $\btheta$ which include standard deviations, first-order autoregressive model correlation parameters, and reparameterised Besag-York-Mollie model [@simpson2017penalising] proportion parameters.

@@

```{r}
time_taken <- read_csv("depends/time-taken.csv")
```

@@

We increased the chain lengths until all NUTS diagnostics were acceptable.
This required chains of length 20,000, which we thinned by a factor of 20 for ease-of-storage.
All potential scale reduction factors [@gelman1992inference; @vehtari2021rank] were $\hat R < 1.05$.
For full details see Appendix S2.
Though inaccuracies remain possible, we considered the NUTS results to be a gold-standard.

@@

```{r node-positions, fig.cap="The 6561 PCA-AGHQ node positions (green, rug plot) projected onto the hyperparameter marginal posteriors (grey, histogram) for each of the 24 hyperparameters. Some hyperparameters, such as logit\\_phi\\_alpha\\_x, are well covered where as others, such as log\\_sigma\\_lambda\\_x, are near only being covered by one unique node."}
knitr::include_graphics("depends/nodes-samples-comparison.png")

# This plot could be improved with a scale for "coverage"
```

@@

The ordered eigenvectors can be interpreted to inform us about the directions of greatest variation in the inverse curvature (Figure S6)
The first and second eigenvectors are the weakly identified AR1 standard deviation and correlation parameters.
Variability can be explained both by low correlation or by high standard deviation.

@@

Let $\phi$ be a generic model parameter.
To assess the informativeness of the data we compared the prior variance $\sigma_\text{prior}^2(\phi)$ to the posterior variance $\sigma_\text{posterior}^2(\phi)$ via the posterior contraction $c(\phi) = 1 - (\sigma_\text{posterior}^2(\phi) / \sigma_\text{prior}^2(\phi))$ [@schad2021toward].
We found that (Figure \ref{fig:contraction})) something something.
For greater interpretability, facet parameters in this plot according to model component.

@@

We assessed the coverage of our estimates via the uniformity of the data within each posterior marginal distribution.
Let $\{\phi_i\}_{i = 1}^n$ be posterior marginal samples.

@@

The two-sample Kolmogorov-Smirnov (KS) test statistic [@smirnov1948table] is the maximum absolute difference between two ECDFs $F(\varphi) = \frac{1}{n} \sum_{i = 1}^n \mathbb{I}_{\phi_i \leq \varphi}$.
See an illustration of the KS test in Figure \ref{fig:ks} and a summary of the results in Table.

@@

Let $\{\bphi_i\}_{i = 1}^n$ be joint posterior samples.
Pareto-smoothed importance sampling [PSIS; @vehtari2015pareto, @yao2018yes] is a method for stabilising the ratios used in importance sampling.
Results for the PSIS analysis are pending.

@@

Let $\Phi = \{\bphi_i\}_{i = 1}^n$ and $\Psi = \{\bpsi_i\}_{i = 1}^n$ be two sets of joint posterior samples, and $k$ be a kernel.
The maximum mean discrepancy [MMD; @gretton2006kernel] can be empirically estimated by
\begin{equation*}
\text{MMD}(\Phi, \Psi) = \sqrt{\frac{1}{n^2} \sum_{i, j = 1}^n k(\bphi_i, \bphi_j) - \frac{2}{n^2} \sum_{i, j = 1}^n k(\bphi_i, \bpsi_j) + \frac{1}{n^2} \sum_{i, j = 1}^n k(\bpsi_i, \bpsi_j)}.
\end{equation*}
We set $k(\bphi_i, \bphi_j) = \exp(-\sigma \lVert \bphi_i - \bphi_j \rVert^2)$ with $\sigma$ estimated from data using the `kernlab` \textsc{R} package [@karatzoglou2019package].
As compared with NUTS, the MMD from PCA-AGHQ (0.071) was 11% smaller than that of TMB (0.080).

>>>>

In all, Naomi is a joint model on the observations $\mathbf{y} = (y^{\theta}_I)$ for $\theta \in \{\rho, \alpha, \kappa, \rho^\text{ANC}, \alpha^\text{ANC}, N^\text{ART}\}$ and $I \in \mathcal{I}$.
The structured additive predictors contain intercept effects, age random effects, and spatial random effects which we collectively describe as the latent field $\x$.
The latent field is controlled by hyperparamters $\btheta$ which include standard deviations, first-order autoregressive model correlation parameters, and reparameterised Besag-York-Mollie model [BYM2; @simpson2017penalising] proportion parameters.

@@

```{r}
#' It's a shame here that I have to put "seconds", "hours" and "days" manually and it's not stored in the time taken
time_taken <- read_csv("depends/time-taken.csv")
```

@@

We increased the chain lengths until all NUTS diagnostics were acceptable.
This required chains of length 20,000, which we thinned by a factor of 20 for ease-of-storage.
All potential scale reduction factors [@gelman1992inference; @vehtari2021rank] were $\hat R < 1.05$ (Figure S7).
For full details see Appendix S2.
Though inaccuracies remain possible, we considered the NUTS results to be a gold-standard.

@@

```{r node-positions, fig.cap="The 6561 PCA-AGHQ node positions (green, rug plot) projected onto the hyperparameter marginal posteriors (grey, histogram) for each of the 24 hyperparameters. Some hyperparameters, such as logit\\_phi\\_alpha\\_x, are well covered where as others, such as log\\_sigma\\_lambda\\_x, are near only being covered by one unique node."}
knitr::include_graphics("depends/nodes-samples-comparison.png")
```

@@

```{r}
cor_df <- read_csv("depends/ar1-bym2-cor.csv")

summary_cor_df <- cor_df %>%
  group_by(type) %>%
  summarise(mean_cor = mean(abs(cor)))

ar1_mean_cor <- signif(filter(summary_cor_df, type == "ar1")$mean_cor, 2)
bym2_mean_cor <- signif(filter(summary_cor_df, type == "bym2")$mean_cor, 2)
```

The ordered eigenvectors correspond to the directions of greatest variation in the inverse curvature.
The first and second eigenvectors each contain coupled AR1 standard deviation and correlation parameters (Figure S6).
As variability can be explained either by high standard deviation and high correlation or low standard deviation and low correlation these parameters are weakly identified.
This is reflected in high correlation in the the NUTS posterior pairs plot (Figure S10, average absolute correlation across all four pairs of `r ar1_mean_cor`).
The BYM2 standard deviation and proportion parameters on the other hand are designed to be orthogonal, and as such did not display posterior correlation (Figure S11, average absolute correlation across all four pairs of `r bym2_mean_cor`) or appear prominently in the eigenvectors.

@@

To assess the informativeness of the data we compared the prior variance $\sigma_\text{prior}^2(\psi)$ to the posterior variance $\sigma_\text{posterior}^2(\psi)$ via the posterior contraction $c(\psi) = 1 - (\sigma_\text{posterior}^2(\psi) / \sigma_\text{prior}^2(\psi))$ [@schad2021toward], where $\psi$ is a model parameter.
We found that (Figure \ref{fig:contraction})) something something.
For greater interpretability, facet parameters in this plot according to model component.

@@

We assessed the coverage of our estimates via the uniformity of the data within each posterior marginal distribution.
Let $\{\psi_i\}_{i = 1}^n$ be posterior marginal samples.

@@

The two-sample Kolmogorov-Smirnov (KS) test statistic [@smirnov1948table] is the maximum absolute difference between two ECDFs $F(\omega) = \frac{1}{n} \sum_{i = 1}^n \mathbb{I}_{\psi_i \leq \omega}$.
See an illustration of the KS test in Figure \ref{fig:ks} and a summary of the results in Table.

@@

Let $\{\bpsi_i\}_{i = 1}^n$ be joint posterior samples.
Pareto-smoothed importance sampling [PSIS; @vehtari2015pareto, @yao2018yes] is a method for stabilising the ratios used in importance sampling.
Results for the PSIS analysis are pending.

@@

Let $\Psi = \{\bpsi_i\}_{i = 1}^n$ and $\Psi = \{\bpsi_i\}_{i = 1}^n$ be two sets of joint posterior samples, and $k$ be a kernel.
The maximum mean discrepancy [MMD; @gretton2006kernel] can be empirically estimated by
\begin{equation*}
\text{MMD}(\Psi, \Psi) = \sqrt{\frac{1}{n^2} \sum_{i, j = 1}^n k(\bpsi_i, \bpsi_j) - \frac{2}{n^2} \sum_{i, j = 1}^n k(\bpsi_i, \bpsi_j) + \frac{1}{n^2} \sum_{i, j = 1}^n k(\bpsi_i, \bpsi_j)}.
\end{equation*}
We set $k(\bpsi_i, \bpsi_j) = \exp(-\sigma \lVert \bpsi_i - \bpsi_j \rVert^2)$ with $\sigma$ estimated from data using the `kernlab` \textsc{R} package [@karatzoglou2019package].
As compared with NUTS, the MMD from PCA-AGHQ (0.071) was 11% smaller than that of TMB (0.080).
